---
title: "HABS-HD Summaries"
author: "ATRI Biostatistics"
output:   
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib  
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{HABS-HD Summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
library(HABSHD)
library(arsenal)
library(ggplot2)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, 
  fig.height = 8 / 1.6, 
  out.width = "100%"
)

options(digits = 2, width = 150)

theme_set(theme_minimal())

scale_colour_discrete <-
    function(...) ggsci::scale_color_lancet()
scale_fill_discrete <-
    function(...) ggsci::scale_fill_lancet()
```

The **The Health & Aging Brain Study: Health Disparities (HABS-HD)** study [@OBryant2021health] is a single site project that is conducted at the University of North Texas Health Science Center. The HABS-HD project is funded by grants from the National Institute on Aging (NIA): R01AG054073, R01AG058533. The HABS-HD study has now enrolled over 1,000 Mexican Americans and 1,000 non-Hispanic whites from the community. Enrollment for an additional 1,000 African Americans began February 2021.

The code below demonstrates how to use data in the `HABSHD` package and @R to derive some basic summaries.

```{r, child='../man/rmd/HABS-HD-Summaries-child.Rmd'}
```

# References
